Handling missing data in smartphone location logs

### 2017-11-26 thesisReport.Rmd
simplified introduction

How active people are and how they interact with their environment affects a wide range of outcomes including health, income and social capital [@goodchild_toward_2010]. A better understanding of both within-person and between-person variability in geospatial patterns could be conducive to better social, health and urban-planning policies. Yet a large part of studies on human mobility are largely based on pen-and-paper travel diaries. These surveys have known methodological flaws, such as the short period of data collection (due to costs and burden to respondents), the underreporting of short trips [@wolf_impact_2003] and the underestimation of the duration of commutes [@delclos-alio_keeping_2017].

These obstacles can be overcome by using objective data on human mobility. The Global Positioning System (GPS), which uses the distance between a device and a number of satellites to determine location, provides such data. Within behavioural science, this type of data has been used to investigate topics such as the effects of the food environment on eating patterns [@zenk_how_2009], the movement correlates of personality [@harari_using_2016], academic performance [@wang_smartgpa:_2015] and bipolar disorder [@palmius_detecting_2017].

In most of these studies participants are given a specialised device, resulting in accurate mobility GPS data (*specialised logs*). However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants. Moreover, they may be biased because of the introduction of a new device to the participant's life. Because of these drawbacks, specialised logs usually span a short amount of time. @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*). Another solution is to take advantage of existing smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years [@location_history_timeline_2017] (*secondary logs*). By law, logs can be accessed and shared by users for free [@commission_protecting_2017]. Yet, because they were created for non-academic purposes under engineering constraints, the sensors do not monitor continuously and the resulting logs can be sparse and inaccurate. Hence, two important challenges are dealing with measurement noise and missing data.

Missing data is a pervasive issue as it can arise due to multiple reasons. Technical reasons include signal loss, battery failure and device failure. Behavioural reasons include leaving the phone at home, switching the phone off, switching location measurements off, and so on. As a result, applied researchers are often left with wide temporal gaps with no measurements. For instance, different groups studying the effect of bipolar disorder on human movement have reported missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017]. Similar trends are consistently reported in other fields [e.g. @harari_using_2016;@jankowska_framework_2015].

There is currently no golden standard in how to deal with missing data in custom or secondary logs [@barnett_inferring_2016]. Traditional missing data methods, such as mean imputation, cannot be used easily in spatiotemporal data because the measurements are correlated in time and space. For example, assume the simplistic case that an individual spends almost half her time at work, half her time at home and a small amount of time commuting between the two along a angled path. Using mean imputation would result in imputed values of her being at the geometric midpoint between the home and work for all missing values, even though she has never been there and never will be. Worringly, @jankowska_framework_2015 have pointed out that there is often little transparency regarding decisions of how to deal with missing data.

The accuracy of GPS measurements in smartphones is substantially lower than in professional grade GPS trackers. For example, Android phones collect location information through a variety of methods, such as from WiFi access points, cellphone triangulation, and GPS measurements. They use different methods due to computational and battery constraints [@lamarca_place_2005; @chen_practical_2006].In professional grade GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are reported to be most inaccurate in high density urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for social scientists, this happens to be where most people in the developed world tend to spend most of their time.

On the other hand, noisy data can lead to inaccurate conclusions if it is not accounted for, such as overestimating the movement of individuals. For instance, suppose that a naive researcher calculates the distance travelled by an individual by drawing a line between each measured point and calculating the sum of the length of all of these lines. If there is noise, the measurements will vary even though the individual is not moving. If the measurements are frequent, then the researcher will end up with a lot of movement, even though the individual did not move at all. The problem is further complicated by the fact that missing data and noisy measurements are related to each other. For instance, methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017] can exacerbate the severity of the missing data problem.

@@

In the example data set the distribution of $a$ is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km. @palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. We observe similar patterns in secondary logs. Figure 1 shows how accuracy can vary as a function of user behaviour, time and location. Inaccurate measures are often followed by more accurate measures.  Most notably, low accuracy tends to be associated with movement (Figure 2). Stationary accuracy varies depending on phone battery level, wifi connection and user phone use. There are several recurring low-accuracy points, possibly the result of cell-phone tower triangulation.

>>>>

How people move about in their environment affects a wide range of outcomes. These include health, income and social capital [@goodchild_toward_2010]. A better understanding of mobility could lead to better health and urban-planning policies. A large part of studies on human mobility are conducted with pen-and-paper travel diaries. These surveys have known methodological flaws. The high cost and burden to respondents limits the period of data collection. Short trips are often under reported  [@wolf_impact_2003] . The self-reported duration of commutes is often underestimated [@delclos-alio_keeping_2017].

These obstacles can be overcome by using objective data on human mobility. Such data  can now be obtained using the Global Positioning System (GPS). GPS uses the distance between a device and several satellites to determine location. Within behavioural sciences, researchers have used GPS data to investigate wideranging topics. For example,  @zenk_how_2009 investigate the effects of the food environment on eating patterns. @harari_using_2016 look at the movement correlates of personality.  @wang_smartgpa:_2015  does the same for academic performance.  @palmius_detecting_2017 use mobility patterns to predict bipolar depression.

In most studies participants receive a specialised GPS devices to track their mobility.  We call resulting logs *specialised logs*. @barnett_inferring_2016 point out several methodological issues with these studies. High costs and a high burden to participants limits their scalability. Besides, introducing a new device to the participant's life may bias the data. Due to these drawbacks specialised logs usually span a short amount of time. @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*). Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) . For instance,  Google Location History contains information on millions of users [@location_history_timeline_2017]. Often, secondary logs span several years. By law, secondary logs are accessible to users for free [@commission_protecting_2017]. Yet, secondary logs also present methodological challenges. They were created for non-academic purposes under engineering constraints. These constraints mean that sensors do not track users continuously. The resulting logs can be sparse and inaccurate. Hence, two important challenges are dealing with measurement noise and missing data.

Missing data is a pervasive issue as it can arise due to several reasons. Technical reasons include signal loss, battery failure and device failure. Behavioural reasons include leaving the phone at home or switching the device off. As a result, secondary logs often contain wide temporal gaps with no measurements. For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017]. Other researchers report similar trends in other fields [e.g. @harari_using_2016;@jankowska_framework_2015].

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Traditional methods, such as mean imputation, are not suitable for spatiotemporal data. This is because measurements are often correlated in time and space. For example, imagine an individual who splits almost all her time between work or at home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work.  She has never and will never be there! Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

Smartphone location measurements are substantially less accurate than those of professional GPS trackers. Android phones collect location information through a variety of less accurate methods. Other than GPS measurements, Androids use WiFi access points and cellphone triangulation. Different methods are used because of computational and battery constraints [@lamarca_place_2005; @chen_practical_2006]. In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time.

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. If the measurements are frequent and noisy, we will calculate a lot of movement. Even if the individual did not move at all! The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

@@

In the example data set the distribution of $a$ is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km. @palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. We observe similar patterns in secondary logs. Figure 1 shows how accuracy can vary as a function of user behaviour, time and location. Inaccurate measures are often followed by more accurate measures.  Most notably, low accuracy often (but not always) is  associated with movement (Figure 2). Stationary accuracy varies depending on phone battery level, wifi connection and user phone use. There are several recurring low-accuracy points, possibly the result of cell-phone tower triangulation.
